Cosmic web alignments with the shape, angular momentum and peculiar velocities of dark matter halos

### 2014-03-13 paper/main.tex
many new references

We study the alignment of dark matter halos with the cosmic web as described by
the tidal and velocity shear fields. We focus on the alignment of their
shape, angular momentum and peculiar velocities. We use a cosmological N-body
simulation that allows to study dark matter halos spanning almost five orders
of magnitude in mass ($10^{9}$-$10^{14}$) \hMsun and use spatial scales
of $(0.5$-$1.0)$\hMpc to define the cosmic web. After varying the
numerical parameters in our experiments we find that the halo shape
and the tidal field web presents the strongest alignment along
filaments and walls with a signal that gets stronger as the halo mass
increases; in the case of the velocity shear only massive halos
$>10^{12}$\hMsun tend to have their shapes perpendicular to
filaments and walls. For the angular momentum we find alignment
signals only for halos more massive than $10^{12}$\hMsun both in the
tidal and velocity shear webs where the preferences are to be
perpendicular to filaments or parallel to walls,
respectively. Finally, the peculiar velocities show a strong alignment
for all halo masses to move along filaments and walls in the tidal
web, while in the velocity shear the same alignment is present but
weaker and only for halos less massive than $10^{12}$\hMsun. Most of
the different signals between the two web description are due to the
fact that the main directions of stronger tidal tension and
velocity shear coincide for halos below $10^{12}$\hMpc but are
perpendicular to each other for halos more massive than this
threshold.  This clearly separates halos into those that live
in sheets and filaments (lower masses) and nodes of the cosmic web.

\end{abstract}
\begin{keywords}
methods: N-body simulations, galaxies: haloes, cosmology: theory, dark
matter, large-scale structure of Universe 
\end{keywords}

@@

There is a long observational tradition stuying galactic properties
as a function of their large scale environment. In these situations the environment
definition is usually based on quantities easily accesible to observations
such as the number density. With the advent
of large
surveys the visibility of the cosmic web is clear. As a consequence,
in order to capture this filamentarity, the environment definition
started to be more complex, involving shear properties of the galaxy
density field or the reconstructuted dark matter density field.

The emergence of this web like structure is a great success of large
scale structure growth models based on gravitational
instability. 
Numerical simulations allow the time description
of such structures together with the formation of the virialized dark
matter halos, which in turn should host observable galaxies. The
finding of gas filaments that feed galaxies is another theoretical
hint that strenghtened the expected connection between galactic
properties ant their place in the cosmic web.

With the aid of high resolution simulations it has been established that halos of
a given mass form earlier in denser regions; concentration, spin and
shape also depend on the local density. In the last decade new
algorithms have implemented cosmic web classifications that go beyond
the local density by defining a location to be a peak, sheet, filament
or void depending on the simmetry properties of the local dark matter
distribution.

There are analytical frameworks, such as tidal torque theory, that
aim at describing how galaxies gain angular momentum and expect a
correlation of this quantity with the cosmic web. The observational
studies that try to measure such correlations actually try to use as a
proxy the observed shape of galaxies. In this respect it is useful to
have a theoretical baseline for the correlations of angular momentum
and shape with the cosmic web.

In the same spirit of describing the place of galaxies within the
cosmic web, there has been an revival of surveys that measure the
cosmic flow patterns in the Local Universe. Assuming the correspondence
between the divergence of the cosmic flow velocity field and the local
matter overdensity (valid in the linear regime) one could construct
accurate maps of the matter density from peculiar velocities. From
this perspective it is interesting to look at the expected alignment of
the peculiar velocities.

There is large tradition of alignment measurements of shape and
angular momentum. The main result of these previous studies is that shape
alignment is a robust measurement regardless of the methods and
simulations. On the other hand, the results for the angular momentum
differ in the degree of the alignment. In this \documentname we review
most of these studyes and offer our own study with complementary
numerical techniques and simulations.  We also present for the first
time in the literature new results for the alignment of peculiar velocities
with large scale structure.

The structure of this \documentname is the following. In \S
\ref{sec:theory} we present the theoretical antecedents for the
alignment studies we present in this paper. In \S
\ref{sec:nbody} we present the N-body
cosmological simulation and halo catalogs.  Next, we describe in
\S\ref{sec:algorithms} the two web-finding algorithms we use. In
\S\ref{sec:results} we present our main results about the alignment
of shape, spin and peculiar velocities with respect to the cosmic
web. We continue with a discussion of these results in
\S\ref{sec:discussion} paying special attention to resolution effects
and the conditions that might drive the strong alignment signals
observed in the precedent section. In \S\ref{sec:conclusions} we
present our conclusions.

@@

@@

@@

We note that we measure alignments along the eigenvectors of the
cosmic web without defining first whether each point corresponds to a
filament or a wall. Our motivation is two-fold. First, partitioning the
simulation into web elements implies a choice as to what value to use
for the parameter $\lambda_{\rm th}$; although, this has been done
before we consider that  COMPLETAR...

@@

\begin{figure*}
\includegraphics[width=0.8\textwidth]{Fig1.pdf}
\caption{Fraction of halos in a mass bin that show a preferential
  alignment with respect to an eigenvector. Each row presents on of
  the three properties studied in this paper: shape (major axis),
  angular momentum and  peculiar velocity. Left column presents the
  results against the Vweb   and the right the Tweb. Strong colors
  refer to $256^3$ grid resolutions and lighter colors to a $512^3$
  grid. The thick black horizontal line at $0.33$ corresponds to the
  expected fraction for a random vector field.
\label{fig:preferential}}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=0.75\textwidth]{Fig2.pdf}
\caption{Median of $\cos\theta$ quantifying the shape alignment for
  the Vweb (left) and the Tweb (right) at two different grid
  resolutions. In the upper (lower) panels the angle $\theta$ is
  measured between the halo major axis and the first (third)
  eigenvector.\label{fig:shape_alignment}.  SUGIERO PONER AMBAS LEYENDAS EN UN SOLO PANEL, SINO PARECE QUE
IZQUIERDA Y DERECHA FUERAN PARA DOS RESOLUCIONES DISTINTAS.} 
\end{figure*}

@@

MOVI EL TITULO DE LA SUBSECCION AQUI, ESTABA DOS PARRAFOS MAS ABAJO, PERO TAMBIEN MODIFIQUE EL TEXTO INICIAL
\subsection{Interweb Alignment}

%We also wish to directly compare the results of the two
%web tensor algorithms. 
For the two algorithms, Tweb and Vweb, we have the information for their
eigenvectors and eigenvalues on exactly the same positions defined
by the grids. This allows us tocompute the pair-wise alignment
between the eigenvectors in the two web finders.

We are mainly interested in the alignment of the web near
halos. Therefore we restrict our analysis to the grid cells that are
occupied by halos. Otherwise, if we decided to perform this kind of analysis
on all the grid cells, the statistics would be dominated by the void
regions, because they dominate in number the fraction of cells in the
simuation.

@@

@@

A much weaker alignment signal is present for the angular momentum. In
the Tweb only the most massive halos $>10^{12}$\hMsun are antialigned
with respect to the filaments, while for the Vweb the massive halos
are alignment with sheets. These results broadly agree with the
published literature. Nevertheless, in the same publications there is an
alignment/anti-alignment signal reported at lower halo masses
$<10^{12}$\hMsun that we do not detect in our measurements.

A new result from our study is the alignment for the peculiar
velocities. Here we find a relatively strong signal of alignment along
the direction defined by the third eigenvector and perpendicular to
the first. This signal is very strong in the Tweb for all masses below
$<10^{13}$\hMsun. This can be interpreted as a flow parallel to walls
and filaments. In the case of the Vweb similar signal, albeit weaker,
is present only for the low mass halos $<10^{12}$\hMsun.

@@

@@

The tidal connection between voids in simulations \citep{Platen2008},
the alignment of observed galaxies with the inferred tidal
field\citep{Lee2007,Jones2010}. There are other kind of alignment
statistics based on modifications of the correlation
\citep{Paz2008,Faltenbacher2009} that go beyond a local computation
and therefore are not reviewed in detail in this paper.

@@

\item
\cite{Libeskind2013} %various alignments

@@

\item 
\citep{AragonCalvo2007} %angular momentum alignment

@@

\citep{Hahn2007}% various alignments with the Tweb

>>>>

We study the alignment of dark matter halos with the cosmic web as described by
the tidal and velocity shear fields. We focus on the alignment of their
shape, angular momentum and peculiar velocities. We use a cosmological N-body
simulation that allows to study dark matter halos spanning almost five orders
of magnitude in mass ($10^{9}$-$10^{14}$) \hMsun and use spatial scales
of $(0.5$-$1.0)$\hMpc to define the cosmic web. After varying the
numerical parameters in our experiments we find that the halo shape
and the tidal field web presents the strongest alignment along
filaments and walls with a signal that gets stronger as the halo mass
increases; in the case of the velocity shear only massive halos
$>10^{12}$\hMsun tend to have their shapes perpendicular to
filaments and walls. For the angular momentum we find alignment
signals only for halos more massive than $10^{12}$\hMsun both in the
tidal and velocity shear webs where the preferences are to be
perpendicular to filaments or parallel to walls,
respectively. Finally, the peculiar velocities show a strong alignment
for all halo masses to move along filaments and walls in the tidal
web, while in the velocity shear the same alignment is present but
weaker and only for halos less massive than $10^{12}$\hMsun. Most of
the different signals between the two web description are due to the
fact that the main directions of stronger tidal tension and
velocity shear coincide for halos below $10^{12}$\hMpc but are
perpendicular to each other for halos more massive than this
threshold.  This clearly separates halos into those that live
in sheets and filaments (lower masses) and nodes of the cosmic web. As
a general conclusion, our results clearly show that the two different
algorithms we used to define the cosmic web describe different physical
aspects of non-linear collapse and should be used in a complementary
way according to the problem at hand.
\end{abstract}
\begin{keywords}
methods: N-body simulations, galaxies: haloes, cosmology: theory, dark
matter, large-scale structure of Universe 
\end{keywords}

@@

There is a long observational tradition stuying galactic properties
as a function of their large scale environment
\citep[e.g.][]{Oemler1974,Dressler1980,Pimbblet2002,Gomez2003,Kauffmann2004,Abbas2006,Baldry2006,Park2007,OMill2008,Wilman2010}. In
these situations the environment definition is usually based on
quantities easily accesible to observations such as local number
density or nearest neighbour measurements \citep{Muldrew2012}. With
the advent of large surveys the visibility of the cosmic web is
clear. As a consequence, in order to capture this filamentarity, the
environment definition started to be more complex, involving shear
properties of the galaxy density field or the reconstructuted dark
matter density field \citep[e.g.]{Lee2005,Sousbie2008,MunozCuartas2011,Tempel2014}.

In parallel, numerical simulations can reproduce the web-like
structure of the galaxy distribution in models based on gravitational
instability in a Dark Matter (DM) dominated universe
\citep[e.g.][]{Bond1996,Colberg2005}. Simulations now allow to follow the time
evolution into the non-linear regime of virialized structures which in
turm should host observable galaxies. The discovery in simulations of
gas filaments that feed galaxies is another theoretical hint that
strenghtened the expected connection between galactic properties ant
their place in the cosmic web \citep{Ocvirk2008,Dekel2009}.

With the aid of high resolution simulations it has been established that halos of
a given mass form earlier in denser regions; concentration, spin and
shape also depend on the local density. In the last decade new
algorithms have implemented cosmic web classifications that go beyond
the local density by defining a location to be a peak, sheet, filament
or void depending on the simmetry properties of the local dark matter
distribution
\citep[e.g][]{Hahn2007,Sousbie2008b,Tweb,Gonzalez2010,Noh2011}.

There are analytical frameworks, such as tidal torque theory, that
aim at describing how galaxies gain angular momentum and expect a
correlation of this quantity with the cosmic web. The observational
studies that try to measure such correlations actually try to use as a
proxy the observed shape of galaxies
\citep{Lee2002,Lee2007,Jones2010}. In this respect it is useful to
have a theoretical baseline for the correlations of angular momentum
and shape with the cosmic web.  There is large tradition of alignment
measurements of shape and angular momentum
\citep[e.g.][]{Hahn2007,AragonCalvo2007,Zhang2009,Codis2012,Trowland2013,Libeskind2013}. The
main result of these previous studies is that shape alignment is a
robust measurement regardless of the methods and simulations. On the
other hand, the results for the angular momentum differ in the degree of the alignment.

In the same spirit of describing the place of galaxies within the
cosmic web, there has been an revival of surveys that measure the
cosmic flow patterns in the Local Universe
\citep{Nusser2011,Tully2013}. Assuming the correspondence between the
divergence of the cosmic flow velocity field and the local matter
overdensity (valid in the linear regime) one could construct accurate
maps of the matter density from peculiar velocities
\citep{Courtois2012}. From this perspective it is interesting to look
at the expected alignment of the peculiar velocities.

In this \documentname we review most of these studies about shape and
angular momentum alignment and offer our own study with complementary
numerical techniques and simulations.  We also present for the first
time in the literature new results for the alignment of peculiar velocities
with large scale structure.

The structure of this \documentname is the following. In \S
\ref{sec:theory} we present the theoretical antecedents for the
alignment studies we present in this paper. In \S
\ref{sec:nbody} we present the N-body
cosmological simulation and halo catalogs.  Next, we describe in
\S\ref{sec:algorithms} the two web-finding algorithms we use. In
\S\ref{sec:results} we present our main results about the alignment
of shape, spin and peculiar velocities with respect to the cosmic
web. In \S\ref{sec:conclusions} we present our conclusions.

@@

In both algorithms we note the three eigenvalues as
$\lambda_1>\lambda_2>\lambda_3$ and their correspoding eivenvectors as
$\hat{e}_1$, $\hat{e}_2$ and $\hat{e}_3$.

@@

We caution the reader that the results reported by
\citep{Vweb, Libeskind2013} use a velocity field that is calculated by a
gaussian smoothing of the CIC velocity field without taking into
account any weight by mass. We present a simple example to clarify the difference
between the two ways of finding the smoothed velocity field. Let's consider two
neighboring CIC cells, one with total mass $m_1 = 10^{11}$\hMsun and
velocity along $x$ $v_{1x}=0$ km/s and a second cell with mass $m_2 = 10^{10}$
and $v_{2x}=-10$ km/s. In the case of defining an average value for the
velocity over these two cells our method finds a result $\bar{v}_x=
(m_1v_{1x} + m_2v_{2x})/(m_1+m_2)= -0.09$km/s while the method used in
\cite{Vweb,Libeskind2013} would find
$\bar{v}_x=(v_{1x}+v_{2x})/2=-5$km/s.

@@

We note that we measure alignments along the eigenvectors of the
cosmic web without defining first whether each point corresponds to a
filament or a wall. Our motivation is two-fold. First, partitioning the
simulation into web elements implies a choice as to what value to use
for the parameter $\lambda_{\rm th}$; although, this has been done
before we consider that deriving robust results independent of a the
free paramenter $\lambda_{\rm th}$ can be more illuminating to
understand the connection of halos with the cosmic web.

However, for simplicity and readibility we report our results in terms
of alignment with respect to filaments and sheets. Given that the
direction along a filament should be defined by the eigenvector
$\hat{e}_3$ corresponding to the smallest eigenvalue $\lambda_3$, a
strong aligmnet along that vector will be reported as an aligment
along filaments. Correspondingly,

@@

\begin{figure*}
\includegraphics[width=0.8\textwidth]{Fig1.pdf}
\caption{Fraction of halos in a mass bin that show a preferential
  alignment with respect to an eigenvector. Each row presents on of
  the three properties studied in this paper: shape (major axis),
  angular momentum and  peculiar velocity. Left (right) column presents the
  results against the Vweb (Tweb). Strong colors refer to $256^3$ grid
  resolutions and lighter colors to a $512^3$ grid. The thick black
  horizontal line at $0.33$ corresponds to the expected fraction for a
  random vector field. 
\label{fig:preferential}}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=0.75\textwidth]{Fig2.pdf}
\caption{Median of $\cos\theta$ quantifying the shape alignment for
  the Vweb (left) and the Tweb (right) at two different grid
  resolutions. In the upper (lower) panels the angle $\theta$ is
  measured between the halo major axis and the first (third)
  eigenvector.\label{fig:shape_alignment}. }
\end{figure*}

@@

@@

\subsection{Interweb Alignment}

For the two algorithms, Tweb and Vweb, we have the information for their
eigenvectors and eigenvalues on exactly the same positions defined
by the grids. This allows us tocompute the pair-wise alignment
between the eigenvectors in the two web finders.

We restrict our analysis to the grid cells that are occupied by
halos. Otherwise, if we decided to perform this kind of analysis on
all the grid cells, the statistics would be dominated by the void
regions as they dominate in number the fraction of cells in the
simuation.

@@

A much weaker alignment signal is present for the angular momentum. In
the Tweb only the most massive halos $>10^{12}$\hMsun are antialigned
with respect to the filaments, while for the Vweb the massive halos
are alignment with sheets. These results broadly agree with the
published literature. Nevertheless, in some publications
\citep{AragonCalvo2007,Hahn2007} there is an alignment signal
reported at lower halo masses $<10^{12}$\hMsun that we do not detect
in our measurements.

A new result from our study is the alignment for the peculiar
velocities. Here we find a relatively strong signal of alignment along
the direction defined by the third eigenvector and perpendicular to
the first. This signal is clear in the Tweb for all masses below
$<10^{13}$\hMsun. This can be interpreted as a flow parallel to walls
and filaments. In the case of the Vweb similar signal, albeit weaker,
is present only for the low mass halos $<10^{12}$\hMsun.

@@

Our study has shown that the alignment properties of dark matter
halos depend on the physical definition of the cosmic web. This shows
that the main aspect of non-linear gravitational collapse might be
revealead easily depending on the choice of the physical context
(i.e. tidal field vs. velocity shear). There is not a better method,
simply differente perspectives. We will explore the potential of these
two different techniques in undertanding the environmental dependence
of galaxy evolution in future work.

@@

In this Appendix we review the results that use a similar exploration
techniques for halo alignments with the cosmic web. Other kind of
alignment statistics based based on modifications of the correlation
\citep{Paz2008,Faltenbacher2009} that go beyond a local computation
and therefore are not reviewed here.

@@

\item
\cite{Libeskind2013}

@@

\item 
\citep{AragonCalvo2007}

@@

\citep{Hahn2007}
